---
title: "Simulation, Fitting, and Bootstrap with Log Link"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation, Fitting, and Bootstrap with Log Link}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5
)
```

```{r setup}
library(SensIAT)
library(dplyr)
library(ggplot2)
library(splines)
set.seed(12345)
```

# Introduction

This vignette demonstrates a complete workflow for sensitivity analysis with the SensIAT package using a **log link function**. We will:

1. **Simulate** longitudinal data with irregular assessment times
2. **Fit** a marginal mean model using the log link
3. **Perform bootstrap resampling** to obtain confidence intervals
4. **Visualize** the results and compare across sensitivity parameters

This workflow is particularly useful for outcome variables that must remain positive (such as counts, concentrations, or other strictly positive measurements).

# Step 1: Simulate Data with Log Link

We start by simulating a realistic longitudinal dataset where:
- The outcome follows a log-normal distribution
- The outcome model incorporates the previous outcome, time effects, and inter-visit spacing
- Observation times are generated through an intensity process dependent on the previous outcome

```{r simulate}
# Set parameters for data generation
n_subjects <- 200
end_time <- 830  # ~2.3 years in days

# Define outcome model coefficients for the log link
outcome_coef <- list(
    prev_outcome = c(0.5, -0.08, 0.02),  # Natural spline coefficients for log(previous outcome)
    time = -0.0008,                        # Trend over time
    delta_time = -0.0015,                  # Effect of time since last visit
    intercept = 0.8                        # Intercept on log scale
)

# Simulate the dataset
sim_data <- simulate_SensIAT_data(
    n_subjects = n_subjects,
    End = end_time,
    intensity_coef = -0.05,
    outcome_coef = outcome_coef,
    baseline_hazard = 0.005,
    outcome_sd = 0.4,
    seed = 42,
    link = "log"  # Specify log link for outcome simulation
)

# Display summary of simulated data
cat("Number of subjects:", n_distinct(sim_data, Subject_ID), "\n")
cat("Number of observations:", nrow(sim_data), "\n")
cat("Outcome range:", range(sim_data$Outcome), "\n")
cat("Time range:", range(sim_data$Time), "\n")

head(sim_data, 10)
```

## Explore Simulated Data

Let's visualize a sample of subjects to see the irregular assessment patterns:

```{r plot_sim_data}
# Plot outcomes for a random sample of subjects
sample_subjects <- sample(unique(sim_data$Subject_ID), size = 12, replace = FALSE)

sim_data %>%
    filter(Subject_ID %in% sample_subjects) %>%
    ggplot(aes(x = Time, y = Outcome, group = Subject_ID)) +
    geom_line(alpha = 0.5, color = "steelblue") +
    geom_point(alpha = 0.6, color = "steelblue", size = 2) +
    facet_wrap(~Subject_ID, ncol = 4) +
    labs(
        title = "Observed Outcomes over Time",
        subtitle = "Sample of 12 subjects with irregular assessment times",
        x = "Time (days)",
        y = "Outcome (log scale)",
        caption = "Simulated data with log link"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Step 2: Fit Model with Log Link

Now we fit a `SensIAT_within_group_model` using the log link function. The log link is particularly suitable when:
- The outcome is strictly positive
- A multiplicative model structure is more natural than additive
- You want to model relative changes in the outcome

```{r fit_model}
# Define knots for the spline basis
knots <- c(200, 400, 600)

# Fit the model with log link
model_log <- fit_SensIAT_within_group_model(
    group.data = sim_data,
    outcome_modeler = fit_SensIAT_single_index_fixed_coef_model,
    outcome.args = list(
        model = ~ ns(..prev_outcome.., df = 3) + scale(..delta_time..) - 1
    ),
    alpha = c(-0.6, -0.3, 0, 0.3, 0.6),  # Sensitivity parameters
    id = Subject_ID,
    outcome = Outcome,
    time = Time,
    End = end_time,
    knots = knots,
    link = "log",          # Specify log link
    loss = "lp_mse"        # Loss function in transformed space
)

# Display model summary
print(model_log)
```

## Extract and Compare Coefficients

The fitted model provides coefficient estimates for different sensitivity parameters (alpha values):

```{r extract_coefs}
# Extract coefficients for each alpha value
coef_matrix <- coef(model_log)
print(coef_matrix)

# Visualize how coefficients vary with alpha
as.data.frame(coef_matrix) %>%
    rownames_to_column("Alpha") %>%
    mutate(Alpha = as.numeric(Alpha)) %>%
    pivot_longer(cols = -Alpha, names_to = "Coefficient", values_to = "Value") %>%
    ggplot(aes(x = Alpha, y = Value, color = Coefficient)) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    facet_wrap(~Coefficient, scales = "free_y") +
    labs(
        title = "Sensitivity Parameter Estimates",
        subtitle = "Log link model: Coefficient variation across alpha values",
        x = "Sensitivity Parameter (alpha)",
        y = "Coefficient Value"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
```

# Step 3: Generate Predictions

Using the fitted model, we can generate predictions of the marginal mean function:

```{r predictions}
# Create a grid of time points for prediction
pred_times <- seq(0, end_time, by = 50)

# Generate predictions for alpha = 0 (primary analysis)
predictions <- compute_SensIAT_expected_values(
    model_log,
    times = pred_times,
    alpha = 0  # Main sensitivity parameter
)

# Visualize predictions
predictions %>%
    mutate(alpha = as.factor(alpha)) %>%
    ggplot(aes(x = time, y = estimate, fill = alpha, color = alpha)) +
    geom_line(size = 1) +
    labs(
        title = "Estimated Marginal Mean Function",
        subtitle = "Log link model: Mean outcome trajectory",
        x = "Time (days)",
        y = "Expected Outcome",
        color = "Sensitivity\nParameter",
        fill = "Sensitivity\nParameter"
    ) +
    theme_minimal() +
    theme(legend.position = "right")
```

# Step 4: Parametric Bootstrap

To obtain confidence intervals and assess uncertainty in our estimates, we perform parametric bootstrap resampling. This involves:
1. Sampling new coefficient values from the asymptotic MVN distribution of the ML estimates
2. Simulating new datasets under the fitted model
3. Re-fitting the model to each bootstrap sample
4. Computing the distribution of estimates across bootstrap replicates

## Configure Bootstrap Parameters

```{r bootstrap_setup}
# Set up bootstrap control parameters
n_bootstrap <- 100  # Number of bootstrap replicates

# Control parameters for bootstrap
boot_control <- pcori_control(
    compute_influence = TRUE,
    influence_args = list(
        method = "fast"      # Use fast integration method
    )
)
```

## Perform Bootstrap Resampling

```{r bootstrap_resample, eval=FALSE}
# Perform parametric bootstrap
# Note: This step may take several minutes depending on n_bootstrap
bootstrap_results <- parametric_bootstrap(
    object = model_log,
    n_boot = n_bootstrap,
    seed = 123,
    control = boot_control,
    quiet = FALSE
)

# Display bootstrap summary
print(bootstrap_results)
```

```{r bootstrap_resample_actual, include=FALSE, eval=TRUE}
# For demonstration, we'll use a smaller bootstrap sample to keep execution time reasonable
n_bootstrap <- 50
bootstrap_results <- parametric_bootstrap(
    object = model_log,
    n_boot = n_bootstrap,
    seed = 123,
    control = boot_control,
    quiet = TRUE
)
```

## Extract Bootstrap Confidence Intervals

```{r bootstrap_ci}
# Extract bootstrap samples
boot_samples <- bootstrap_results$boot_samples  # List of fitted models from bootstrap replicates

# Extract predictions from bootstrap samples
boot_predictions <- lapply(
    boot_samples,
    function(boot_model) {
        compute_SensIAT_expected_values(
            boot_model,
            times = pred_times,
            alpha = 0
        ) %>%
            select(time, estimate) %>%
            rename(pred = estimate)
    }
)

# Combine bootstrap predictions
boot_pred_df <- bind_rows(boot_predictions, .id = "replicate") %>%
    mutate(replicate = as.numeric(replicate))

# Calculate confidence intervals at each time point
ci_summary <- boot_pred_df %>%
    group_by(time) %>%
    summarise(
        mean_estimate = mean(pred, na.rm = TRUE),
        lower_ci = quantile(pred, 0.025, na.rm = TRUE),
        upper_ci = quantile(pred, 0.975, na.rm = TRUE),
        .groups = "drop"
    )

head(ci_summary)
```

## Visualize Bootstrap Results with Confidence Intervals

```{r bootstrap_plot}
# Plot point estimates and 95% confidence intervals
ci_summary %>%
    ggplot(aes(x = time)) +
    geom_ribbon(
        aes(ymin = lower_ci, ymax = upper_ci),
        alpha = 0.25,
        fill = "steelblue"
    ) +
    geom_line(aes(y = mean_estimate), color = "steelblue", size = 1) +
    geom_line(aes(y = lower_ci), color = "steelblue", linetype = "dashed", alpha = 0.5) +
    geom_line(aes(y = upper_ci), color = "steelblue", linetype = "dashed", alpha = 0.5) +
    labs(
        title = "Marginal Mean with 95% Bootstrap Confidence Intervals",
        subtitle = "Log link model with 50 bootstrap replicates",
        x = "Time (days)",
        y = "Expected Outcome",
        caption = "Shaded region shows 95% CI from parametric bootstrap"
    ) +
    theme_minimal() +
    theme(plot.caption = element_text(hjust = 0, size = 9))
```

## Bootstrap Distribution of Estimates

We can also examine how the point estimates vary across bootstrap replicates:

```{r bootstrap_dist}
# Select a few representative time points
time_points <- c(200, 400, 600)

boot_pred_df %>%
    filter(time %in% time_points) %>%
    mutate(time = as.factor(time)) %>%
    ggplot(aes(x = pred, fill = time)) +
    geom_histogram(bins = 15, alpha = 0.6) +
    facet_wrap(~time) +
    labs(
        title = "Bootstrap Distribution of Mean Predictions",
        subtitle = "Distribution across 50 bootstrap replicates",
        x = "Predicted Outcome",
        y = "Frequency",
        fill = "Time (days)"
    ) +
    theme_minimal()
```

# Step 5: Sensitivity Analysis across Alpha Values

One key feature of SensIAT is the ability to conduct sensitivity analysis. We can see how estimates change under different assumptions about the missing data mechanism:

```{r sensitivity_analysis}
# Compute predictions for all alpha values
alpha_values <- c(-0.6, -0.3, 0, 0.3, 0.6)
pred_all_alpha <- lapply(
    alpha_values,
    function(a) {
        compute_SensIAT_expected_values(
            model_log,
            times = pred_times,
            alpha = a
        ) %>%
            mutate(alpha = a)
    }
) %>%
    bind_rows()

# Visualize sensitivity across alpha
pred_all_alpha %>%
    mutate(alpha = as.factor(alpha)) %>%
    ggplot(aes(x = time, y = estimate, color = alpha, linetype = (alpha == 0))) +
    geom_line(size = 1) +
    scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 2), guide = "none") +
    scale_color_brewer(palette = "RdYlBu") +
    labs(
        title = "Sensitivity Analysis: Effect of Alpha Parameter",
        subtitle = "Log link model across different sensitivity parameters",
        x = "Time (days)",
        y = "Expected Outcome",
        color = "Sensitivity\nParameter (α)",
        caption = "Solid line: primary analysis (α=0), Dashed: sensitivity analyses"
    ) +
    theme_minimal() +
    theme(legend.position = "right")
```

# Summary

This vignette demonstrated a complete workflow for sensitivity analysis with the SensIAT package:

1. **Simulation**: Generated realistic longitudinal data using the log link, incorporating irregular assessment times and outcome dynamics
2. **Model Fitting**: Fitted a flexible marginal mean model with the log link function
3. **Bootstrap Inference**: Performed parametric bootstrap to obtain confidence intervals for the marginal mean trajectory
4. **Sensitivity Analysis**: Examined how estimates vary across different sensitivity parameters (alpha values)

The log link is particularly useful for:
- Positive-valued outcomes (concentrations, counts, biological markers)
- Modeling multiplicative effects
- Understanding relative changes rather than absolute differences

## Key Functions Used

- `simulate_SensIAT_data()`: Simulate longitudinal data with irregular assessment times
- `fit_SensIAT_within_group_model()`: Fit marginal mean model with specified link function
- `compute_SensIAT_expected_values()`: Generate predictions from fitted model
- `parametric_bootstrap()`: Perform parametric bootstrap resampling for confidence intervals
- `pcori_control()`: Configure model fitting and bootstrap options

## Next Steps

- Try different outcome models by modifying the `outcome.args` argument
- Experiment with different link functions (`"identity"`, `"logit"`)
- Increase `n_bootstrap` for more stable confidence intervals
- Combine with domain-specific scientific knowledge to select appropriate alpha values for sensitivity analysis
